Simulated observations of young protoplanetary discs

### 2013-01-22 paper_pc.tex

\title[Simulated observations of young protoplanetary disks]{Young protoplanetary disks}
\author[Tom  Douglas, Paola Caselli, Et al.]{Tom Douglas$^{1}$\thanks{E-mail:
pytd@leeds.ac.uk}, Paola Caselli$^{1}$\\
$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK}
\begin{document}

@@

The formation and early evolution of protoplanetary disks around solar-type and low-mass protostars has little observational support, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e.g. Krasnopolsky et al. 2011; Machida et al. 2011; Braiding \& Wardle 2012). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows, thus observations of the young disks, predicted to have sizes of about 100\,AU and masses as large as 10\% the original core mass (e.g. Joos et al. 2012), are challenging. The use of sensitive interferometers is needed to achieve high angular resolution and spatially/spectrally disentangle the various (disk + envelope + outflow) components as well to filter out the extended emission tracing envelope material.

After the pioneer work of, e.g., Chandler et al. (1995), Brown et al. (2000), Looney et al. (2000), recent interferometric observations have discovered compact embedded disks in a sample of Class 0 (Andr\'e et al. 2000) sources, finding masses between 0.4 and  $>$1 M$_{\odot}$ (Jorgensen et al. 2007, 2009; Enoch et al. 2011). A 130\,AU disk was discovered toward a Class 0 source in Perseus, using Jansky Very Large Array (JVLA) observations of NH$_3$ (Choi et al. 2007). Pineda et al. (2012) observed methyl formate with the Atacama Millimeter and sub-millimeter Array (ALMA) and found evidence of rotation toward one of the proto-binary Class 0 sources embedded in IRAS 16293-2422. These observations are consistent with an almost edge-on disk. When ALMA will be completed, it will be finally possible to spatially resolve these young disks and, for the first time, put stringent constraints on theoretical models. As full-operational ALMA is fast approaching, it is important to provide observational predictions based on dynamical models of young protoplanetary disks.

In this paper, we focus our attention to the young self-gravitating disks of Boley \& Durisen (2008), where episodic heating induced by spiral shocks is present, as this may be a good representation of the earliest phases of protoplanetary disks and an alternative to the young "static" disks studied by, e.g. Visser et al. (2009, 2011). As shown by Ilee et al. (2011), the spiral shocks cause desorption of volatiles from dust icy mantles and trigger gas-phase chemical reactions with activation energies, which are inactive at lower temperatures. These processes produce clear chemical signatures of the disk dynamics. With the use of 3D radiative transfer modeling, we perform simulated ALMA observations of the disk studied by Ilee et al. (2011) and identify the best tracers of the physical structure of self-gravitating disks. The physical, chemical and radiative transfer models are described in Sect\,\ref{sec:description_model}. Radiative transfer results are in Sect.\,\ref{sec:model_results}, while ALMA simulated observations are in Sec.\,\ref{sec:alma_predictions}. Discussions and conclusions can be found in Sect.\,\ref{sec:discussion}.

%1. description of what has been done so far on the modelling and radiative transfer of disks (this includes also more evolved disks -- Visser et al., Walsh et al., Aikawa et al. ..)
%2. focus on the young disks (work done by Machida et al., Dapp, Basu \& Kunz 2012, .. for their formation; work by Boley et al. on the physical evolution ; Ilee et al. for chemistry) 
%3. previous attempts to observe these young disks: (a) simulations (Cossins et al. 2010); (b) observations: Fuente et al. 2010 (AB Aur); Jorgensen \& van Dishoeck (H$_2^{18}$O and the HDO/H$_2$O ratio); Pineda et al. (2012); other papers talking about "hot corinos" (Bottinelli et al. ....)

@@

%-describe the envelope structure (pre-stellar core)

The physical model used to simulate the emission from a young protoplanetary disc is a hybrid model obtained from embedding a gravitationally unstable protoplanetary disk  within an envelope with characteristics similar to the well-studied pre-stellar core L1544, as derived by Keto \& Caselli (2010; hereafter KC2010). The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert (Bonnor 1956; Ebert 1957) sphere with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated (Keto et al., in preparation), and it is shown in Fig.\,\ref{fig:l1544_model}.  \newline

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/L1544model_used.png}
 \label{fig:l1544_model}
 \caption{Model of the pre-stellar core L1544 used as the envelope of the young protoplanetary disk in the hybrid model. Showing temperature (red) and dust (green) temperature in kelvin, log number density (blue) in cm$^{-3}$ and inward velocity$\times$100 (cyan) in m$\,$s$^{-1}$. Adapted from Keto \& Caselli (2010) and Keto et al. (2012, in preparation).}
\end{figure}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/rhoT2.png}
 \label{rhoT} 
 \caption{Left: A 3D plot of log number density (m$^{-3}$) showing the spiral structure in the xy plane and scale height of the disc. Right: The 3D temperature structure of the disc; regions cooler than 40 degrees are not shown in 3D, demonstrating the narrow central region containing hot material.}
\end{figure*}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/columnDensities2.png}
 \label{Chemistry} 
 \caption{Column densities of H$_2$CO, OCS and HCO$^+$ used in the disc model. Figure adapted from Ilee at al. (2011)}
\end{figure*}

@@

%-describe the physical structure of the disk (density and temperature) -- one figure showing the physical structure and the kinematics (2-panels figure)
The pre-stellar core structure is maintained down to a radius of 80\,AU, within which the young protoplanetary disk has been plugged in. The disk structure is derived by the hydrodynamic model of Boley(2007) and Boley (2009). The particular model considered here (as well as in Ilee et al. 2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating disc featuring prominent spiral arms. H$_2$ number densities in the disc range from 10$^{10}$-10$^{19}\,\rm{m}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The model is sampled over a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$au in x, y and z. The gas/dust mass ratio was assumed to be 1/100 throughout both sections model and the opacities were given by the model of dust grains with thick icy mantles and 10$^6\,$yr coagulation from Ossenkopf and Henning (1994). THe gas number density and temperature structure are displayed in Fig. XX. \newline

@@

%-describe the chemistry  - refer to Ilee et al. which have been taken as input to the rad transfer (RT) code
Chemical abundances in the disc were taken from Ilee et. al (2011) which followed the changes of chemical abundances of trace particles moving through the disc as it evolved. The abundances of 125 species related by 1334 reactions were followed through the time evolution of the disc. These abundances were interpolated onto a 51$^3$ grid covering the disc with cells of size 2.2$\times$2.2$\times$0.22$\,$au.  Fig. XX shows a sample of column density maps from Ilee et al. (2011), pointing out the different regions traced by H$_2$CO and OCS, which mostly probe the central warm regions, and HCO$^+$, which preferentially trace the outer spiral pattern as in the central region is destroyed by water molecules (see Ilee et al. 2011 for details).   The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disk, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also Caselli et al. 2012). As discussed in the result section (Sec.\,\ref{sec:model_results}), rough guesses have been made based on values measured toward similar object.

@@

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
%-describe the RT used (LIME) 
The radiative transfer program used, LIME (Brinch \&Hogerheijde 2010), calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in the griding section (NOTE: WHERE IS THE GRIDING SECTION?). At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and turbulent velocity are taken from the model. These points are then smoothed by Lloyds algorithm (Lloyd 1982) in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation (NOTE: YOU NEED TO EXPLAIN THIS - MAYBE IN A FOOTNOTE) and it is down these paths that photon propagation is restricted (figures \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value.\newline

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime_points_rz_histo.png}
 \label{points}
 \caption{A histogram of the point distribution throughout the model.}
\end{figure}

In order to construct the grid, points are randomly selected from the volume, then compared against a reference point (NOTE: WHAT IS THE COMPARISON FOR?). Grid points are selected at random in cylindrical co-ordinates, linearly spaced in z and $\phi$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) as a threshold. After selection of random co-ordinates, the hydrogen density and molecular density at the point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use, if not then another r, $\phi$, z co-ordinate is selected. The weighting function and griding functions (NOTE: CAN YOU PLEASE EXPLAIN WHAT A GRIDING FUNCTION IS?) were selected empirically to sample all the scales while ensuring that the majority of points are located into the inner disc (where the density is higher --IS THIS LAST STATEMENT, WHICH I JUST ADDED, CORRECT?). 20\% of these points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radius of the model) in order to stop too many of the selecected points clustering in the high density disc and leaving the envelope undersampled. In addiation to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See figure \ref{points} for example of the points distribution in r, z. \newline

@@

NOTE: YOU NEED TO ADD A LITTLE HISTORY ON WHY YOU SELECTED THESE SPECIES TO DO THE RADIATIVE TRANSFER. CAN YOU PLEASE RESTRUCTURE THIS SECTION ?
%Figures showing the RT results in a few molecules/transitions (CO, HCO$^+$, HCN, OCS, H2CO, NH3 --- maybe H2O,H2$^{18}$O [to try first])
The results of these simulations for the molecules OCS, HCO$^+$, H$_2$CO and isotopologues of CO are shown here, other molecules simulated but not shown include HCN, HNC, HNO, HCS$^+$ and  CS.
For the purpose of simulating observations the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc) and is inclined 30$^\circ$ to edge on. From these observations integrated intensity, intensity weighted velocity and position velocity diagrams through the centre of the model were created.
The simulations done are focused upon frequencies within ALMA band 7, selected to give the best trade off between resolution and sensitivity for early ALMA science.
(Note the moment 1 and 0 maps were created by integrating between -12.5 to -0.5 km$\,$s$^{-1}$ and +0.5 to +12.5 km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope, this can be seen in some PV diagrams as the strong absorption feature at all positions around zero velocity, moment 1 maps are shown with a cutoff of 1/1000 of the peak emission/absorption value)\newline

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/imageCO_3-2_30deg_contSub.png}

 \caption{CO J=3-2 Continuum subtracted mom0}
\end{figure}

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/imageCO_3-2_30deg_mom1.png}

 \caption{CO J=3-2 moment 1 map}
\end{figure}

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/imageCO_3-2_30deg_PV_centre.png}

 \caption{CO J=3-2 PV through y=0''}
\end{figure}

Figures 6,7 and 8 show synthetic images of the CO J=3-2 line at 345.8$\,$GHz. As the upper level for this line is at 33$\,$K above the ground state this transition should be excited through out disc. given the large and uniform nature of the abundance of CO in the disc this line will be optically thick and this is only mapping the outer regions of the disc. In common with most transitions simulated, the line is seen almost exclusively in absorption against the disc continuum. The region of emission in the centre is where there is no continuum background to be absorbed.  Some indications of spiral structure can be seen but are indistinct. ***not sure how useful these are, might be worth cutting them?***\newline

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/imageOCS_28-27_30deg_composite_contSub.png}
 \label{OCS_mom0}
 \caption{OCS 28-27 Continuum subtracted mom0}
\end{figure}

@@

 \caption{HCO$^+$ 1-0 pv through centre}
\end{figure}

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/imageHCOp_3-2_30deg_PV_centre_2.png}
 \label{hcop_pv}
 \caption{HCO$^+$ 3-2 PV through centre with rotation curve from figure \ref{velocity} *** should the velocities in this be reduced by cos(30) as we're not in the plane of rotation? ***}
\end{figure}

Some molecules such as HCO$^+$ trace only the outer regions of the disc (Ilee 2011) and so can be used to look at the extended velocity and physical structure. In these colder less dense regions we see the molecular lines in emission rather than absorption. Figure \ref{hcop_pv} shows the rotation curve of the model as shown in figure \ref{velocity} against the HCO$^+$ J=3-2 line emission. It is clear that from observations such as these the rotation curve of a disc could be reconstructed.\newline
\newline

In all the simulations except HCO$^+$ we see molecular lines in absorption throughout the majority of the disc and in some cases a small amount of emission towards the centre, where the inclination of the disc allows us the view the low density but moderate temperature region inside the inner edge of the spiral structures without any high temperature and density material behind it.\newline

@@

In this paper we have performed radiative transfer simulations of a hybrid model comprising a 0.4$\, M_\odot$ self gravitating disc with radius 64$\,AU$ showing spiral density waves, surrounded by an envelope simulated as a collapsing 10$\,M_\odot$ BE-sphere. CASA simulations show that at a distance of 100$\,$pc both spatial resolution of the spirals in such a disc and extraction of kinematic and temperature from molecular lines are possible in ALMA band 7. Our simulations show that many  molecular species are predominantly seen in absorption towards the centre of  self-gravitating protoplanetary discs. The quiescent nature of the envelope around such discs only obscures them within $\pm\,0.5 km\,s^{-1}$.\newline
Hot, shocked gas can be detected in such a disc without resolving it by observing OCS transitions between 26-27 to 30-29.\newline
One assumption made in this model is that the gas and dust are in thermal equilibrium.If they are not and the dust is significantly cooler than the gas then transitions may not show up in absorption.\newline
The mass of the disc used (0.4$\,\rm{M}_\odot$) is on the high end for early discs (citations for estimate of early class 1 disc masses).\newline 
In the model spirals shock heat the mid-plane layer and the colder more diffuse gas absorbs the continuum from it, this results in molecular lines being seen in absorbtion whenever lines of sight pass through the cold diffuse gas at larger heights on to the hot dense midplane of the disc. Rotation curves can be gathered from these observations even with spiral structure\newline

>>>>

\title[Young protoplanetary discs]{Simulated observations of young protoplanetary discs}
\author[Tom  Douglas, Paola Caselli, Et al.]{Tom Douglas$^{1}$\thanks{E-mail:
pytd@leeds.ac.uk}, Paola Caselli$^{1}$, Aaron Boley$^{2}$, Richard Durisen$^{3}$,\\ Tom Hartquist$^{1}$, Ilee$^{1}$, Rawlings$^{4}$ \\
$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK\\
%$^{2}$Department of Astronomy, University of Florida, 211 Bryant Space Center, PO Box 112055, USA\\
%$^{3}$Department of Astronomy, Indiana University, 727 East 3rd Street, Swain West 319, Bloomington, IN 47405, USA\\
%$^{4}$Department of Physics & Astronomy, University College London, London WC1E 6BT, UK
}
\begin{document}

@@

The formation and early evolution of protoplanetary discs around solar-type and low-mass protostars has little observational support, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e.g. Krasnopolsky et al. 2011; Machida et al. 2011; Braiding \& Wardle 2012). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows, thus observations of the young discs, predicted to have sizes of about 100\,AU and masses as large as 10\% the original core mass (e.g. Joos et al. 2012), are challenging. The use of sensitive interferometers is needed to achieve high angular resolution and spatially/spectrally disentangle the various (disc + envelope + outflow) components as well to filter out the extended emission tracing envelope material.

After the pioneer work of, e.g., Chandler et al. (1995), Brown et al. (2000), Looney et al. (2000), recent interferometric observations have discovered compact embedded discs in a sample of Class 0 (Andr\'e et al. 2000) sources, finding masses between 0.4 and  $>$1 M$_{\odot}$ (Jorgensen et al. 2007, 2009; Enoch et al. 2011). A 130\,AU disc was discovered toward a Class 0 source in Perseus, using Jansky Very Large Array (JVLA) observations of NH$_3$ (Choi et al. 2007). Pineda et al. (2012) observed methyl formate with the Atacama Millimeter and sub-millimeter Array (ALMA) and found evidence of rotation toward one of the proto-binary Class 0 sources embedded in IRAS 16293-2422. These observations are consistent with an almost edge-on disc. Persson et al. (2012) observed H$_2^{18}$O with ALMA toward the same source and found evidence of relatively large ($\ga$100\,K) excitation temperatures, as well as a HDO/H$_2$O abundance ratio close to that measured on Earth's oceans and Jupiter-family comets (Hartogh et al. 2011). When ALMA will be completed, it will be finally possible to spatially resolve these young discs and, for the first time, put stringent constraints on theoretical models. As full-operational ALMA is fast approaching, it is important to provide observational predictions based on dynamical models of young protoplanetary discs.

In this paper, we focus our attention to the young self-gravitating discs of Boley \& Durisen (2008), where episodic heating induced by spiral shocks is present, as this may be a good representation of the earliest phases of protoplanetary discs and an alternative to the young "static" discs studied by, e.g. Visser et al. (2009, 2011). As shown by Ilee et al. (2011), the spiral shocks cause desorption of volatiles from dust icy mantles and trigger gas-phase chemical reactions with activation energies, which are inactive at lower temperatures. These processes produce clear chemical signatures of the disc dynamics. With the use of 3D radiative transfer modeling, we perform simulated ALMA observations of the disc studied by Ilee et al. (2011) and identify the best tracers of the physical structure of self-gravitating discs. The physical, chemical and radiative transfer models are described in Sect\,\ref{sec:description_model}. Radiative transfer results are in Sect.\,\ref{sec:model_results}, while ALMA simulated observations are in Sec.\,\ref{sec:alma_predictions}. Discussions and conclusions can be found in Sect.\,\ref{sec:discussion}.

@@

The physical model used to simulate the emission from a young protoplanetary disc is a hybrid model obtained from embedding a gravitationally unstable protoplanetary disc  within an envelope with characteristics similar to the well-studied pre-stellar core L1544, as derived by Keto \& Caselli (2010; hereafter KC2010). The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert (Bonnor 1956; Ebert 1957) sphere with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated (Keto et al., in preparation), and it is shown in Fig.\,\ref{fig:l1544_model}.  [NOTE: BESIDES THE PROBLEM WITH THE FIGURE NUMBER, YOU NEED TO READJUST FIGURES, AS THIS ONE IS THE FIRST ONE THAT YOU DESCRIBE, SO IT SHOULD BE NUMBER 1] \newline

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/L1544model_used_legend.png}
 \label{fig:l1544_model}
 \caption{[NOTE: YOU NEED TO ADD LABELS AND UNITS TO THE AXES!!] Model of the pre-stellar core L1544 used as the envelope of the young protoplanetary disc in the hybrid model. Showing temperature (red) and dust (green) temperature in kelvin, log number density (blue) in cm$^{-3}$ and inward velocity$\times$100 (cyan) in m$\,$s$^{-1}$. Adapted from Keto \& Caselli (2010) and Keto et al. (2012, in preparation).}
\end{figure}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/rhoT2.png}
 \label{rhoT} 
 \caption{Left: A 3D plot of log number density (cm$^{-3}$) showing the spiral structure in the xy plane and scale height of the disc. Right: The 3D temperature structure of the disc; regions cooler than 40 degrees are not shown in 3D, demonstrating the narrow central region containing hot material.}
\end{figure*}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/columnDensities2.png}
 \label{Chemistry} 
 \caption{Column densities of H$_2$CO, OCS and HCO$^+$ used in the disc model. Figure adapted from Ilee at al. (2011) [SHOULD WE ALSO ADD C18O COLUMN DENSITY AS WE WANT TO DISCUSS THAT LATER ?? ]}
\end{figure*}

@@

%-describe the physical structure of the disc (density and temperature) -- one figure showing the physical structure and the kinematics (2-panels figure)
The pre-stellar core structure is maintained down to a radius of 80\,AU, within which the young protoplanetary disc has been plugged in. The disc structure is derived by the hydrodynamic model of Boley(2007) and Boley (2009). The particular model considered here (as well as in Ilee et al. 2011) is a 0.39$\,\rm{M}_\odot$ self-gravitating disc featuring prominent spiral arms. H$_2$ number densities in the disc range from 10$^{4}$-10$^{13}\,\rm{cm}^{-3}$, and temperatures range from 30-400 K (figure \ref{rhoT}). The dust and gas temperatures in the disc are assumed to be in equilibrium. The model is sampled over a regular grid of size 256$\times$256$\times$64 with spatial resolution of 0.5$\,$au in x, y and z. The gas/dust mass ratio was assumed to be 1/100 throughout both sections model and the opacities were given by the model of dust grains with thick icy mantles and 10$^6\,$yr coagulation from Ossenkopf and Henning (1994). The gas number density and temperature structure are displayed in Fig. XX. \newline

@@

%-describe the chemistry  - refer to Ilee et al. which have been taken as input to the rad transfer (RT) code
Chemical abundances in the disc were taken from Ilee et. al (2011) which followed the changes of chemical abundances of trace particles moving through the disc as it evolved. The abundances of 125 species related by 1334 reactions were followed through the time evolution of the disc. These abundances were interpolated onto a 51$^3$ grid covering the disc with cells of size 2.2$\times$2.2$\times$0.22$\,$au.  From Ilee et al. (2011) we selected the three species which appear to trace different regions of the disc : the inner 30\,AU (OCS), the inner 60\,AU (H$_2$CO) and the region between $\simeq$20 and 60\,AU (HCO$^+$) and one which traces the entirity of the disc (C$^{18}$O). As explained by Ilee et al. (2011),  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also Caselli et al. 2012). As discussed in the result section (Sec.\,\ref{sec:model_results}), rough guesses have been made based on values measured toward similar objects.

@@

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
%-describe the RT used (LIME) 
The radiative transfer program used, LIME (Brinch \&Hogerheijde 2010), calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in the grid-construction section. At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and turbulent velocity are taken from the model. These points are then smoothed by Lloyds algorithm (Lloyd 1982) in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation \footnote{In three dimensions this means that if four points are conneted into a tetrahedron, the sphere circumscribing these four points contains no other points. It can be shown that this connection is unique for a given set of points.} and it is down these paths that photon propagation is restricted (figures \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in Brinch \&Hogerheijde 2010) to exceed 1000 in 99\% of the points. [NOTE: WHY DO WE NEED TO REACH SUCH A HIGH S/N? --- it isnt s/n in the usual sense, it's a measurement of how converged the model is, defined in brinch 2010 as change in populations this iteration / std-dev of change in pops over last 5 iterations] \newline

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime_points_rz_histo.png}
 \label{points}
 \caption{A 2D histogram of the point distribution throughout the model. The disc and envelope can be seen as two separate entities which have to be sampled using different point distributions}
\end{figure}

@@

[NOTE: I SUGGEST TO:

1. SHOW THE MOMENT 0 AND 1 ONLY MAPS OF C18O, OCS, H2CO AND HCO+ + SPECTRA TOWARD THE CENTER AND +/- 30 AU (FOR EACH SPECIES, YOU CAN DO A 3-PANEL FIGURE SHOWING THE TWO MOMENTS AND A PANEL WITH THE 3 SPECTRA AT DIFFERENT LOCATIONS)
+  THE CONTINUUM MAP 
3. SHOW THE SIMULATED ALMA OBSERVATIONS (SAME STYLE AS BEFORE) AND ALSO SHOW A SEPARATE FIGURE WITH THE PV DIAGRAMS OF ALL SPECIES. 
]

[NOTE: THE FOLLOWING PARAGRAPH IS NOT CLEAR AT ALL !!! PLEASE RE-WRITE MAKING SURE THAT THE SENTENCES MAKE SENSE AND THERE ARE NO TYPOS.  PLEASE MAKE THE FIGURES AS SUGGESTED AND THEN RE-WRITE THIS PART.]
Simulated molecules were confined to speices where the chemical abundance in the disc and envelope are known, or have simulated abundances. In addition radative transfer simulations are limited to are limited to molecules where the collisional and Einstein co-efficients are known. Lines for these molecules were simulated with LIME for freqencies 50$\sim$500$\,$GHz, with focus on lines which could be observed in ALMA band 7 and molecules which are not typically seen in outflows, which havent been included in these simulations and could potentially contaminate disc observations. The results presented in this section represent those lines which show dectectable emission/absorption which can be used to trace either spiral structure or rotation.\newline

%Figures showing the RT results in a few molecules/transitions (CO, HCO$^+$, HCN, OCS, H2CO, NH3 --- maybe H2O,H2$^{18}$O [to try first])
The results of these simulations for the molecules OCS, HCO$^+$, H$_2$CO and isotopologues of CO are shown here, other molecules simulated but not shown include HCN, HNC, HNO, HCS$^+$ and  CS.
For the purpose of simulating observations the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc) and is inclined at 30$^\circ$ to edge on. From these observations integrated intensity, intensity weighted velocity and position velocity diagrams through the centre of the model were created.
The simulations done are focused upon frequencies within ALMA band 7, selected to give the best trade off between resolution and sensitivity for early ALMA science {\bf IS BAND 7 STILL THE BEST FOR US TO LOOK AT SEEING AS WE ARE NOW LOOKING AT ALMA FULL CAPABILITY RATHER THAN CYCLE1?}.
(Note the moment 1 and 0 maps were created by integrating between -12.5 to -0.5 km$\,$s$^{-1}$ and +0.5 to +12.5 km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope, this can be seen in some PV diagrams as the strong absorption feature at all positions around zero velocity, moment 1 maps are shown with a cutoff of 1/1000 of the peak emission/absorption value)\newline
%
%\begin{figure}
% \includegraphics[width=84mm]{Figures/sim/imageCO_3-2_30deg_contSub.png}
%
% \caption{CO J=3-2 Continuum subtracted mom0}
%\end{figure}
%
%\begin{figure}
% \includegraphics[width=84mm]{Figures/sim/imageCO_3-2_30deg_mom1.png}
%
% \caption{CO J=3-2 moment 1 map}
 %\end{figure}
%
%\begin{figure}
% \includegraphics[width=84mm]{Figures/sim/imageCO_3-2_30deg_PV_centre.png}
%
% \caption{CO J=3-2 PV through y=0''}
%\end{figure}

%Figures 6,7 and 8 show synthetic images of the CO J=3-2 line at 345.8$\,$GHz. As the upper level for this line is at 33$\,$K above the ground state this transition should be excited through out disc. given the large and uniform nature of the abundance of CO in the disc this line will be optically thick and this is only mapping the outer regions of the disc. In common with most transitions simulated, the line is seen almost exclusively in absorption against the disc continuum. The region of emission in the centre is where there is no continuum background to be absorbed.  Some indications of spiral structure can be seen but are indistinct. {\bf not sure how useful these are, might be worth cutting them?}\newline

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/imageOCS_28-27_30deg_composite_all.png}
 \label{OCS_mom0}
 \caption{OCS 28-27 {\bf Left:} Continuum subtracted integrated intensity map, {\bf Centre:} Continuum subtracted spectra from Y=0'', X=+0.1'' (bottom) X=-0.1'' (centre) and X=0'' (top) spaced by 10$\,$K for clarity, {\bf Right:} Intensity weighted velocity map}
\end{figure*}

@@

 \caption{HCO$^+$ 1-0 pv through centre, with the rotation curve from figure \ref{velocity} for comparison}
\end{figure}

%\begin{figure}
% \includegraphics[width=84mm]{Figures/sim/imageHCOp_3-2_30deg_PV_centre_2.png}
% \label{hcop_pv}
% \caption{HCO$^+$ 3-2 PV through centre with rotation curve from figure \ref{velocity} *** should the velocities in this be reduced by cos(30) as we're not in the plane of rotation? ***}
%\end{figure}

Some molecules such as HCO$^+$ trace only the outer regions of the disc (Ilee 2011) and so can be used to look at the extended velocity and physical structure. In these colder less dense regions we see the molecular lines in emission rather than absorption. Figure \ref{hcop_pv} shows the rotation curve of the model as shown in figure \ref{velocity} against the HCO$^+$ J=1-0 line emission. It is clear that from observations such as these the rotation curve of a disc could be reconstructed.\newline

In all the simulations except HCO$^+$ we see molecular lines in absorption throughout the majority of the disc and in some cases a small amount of emission towards the centre. Molecular lines throughout the plane of the disc show up in absorption against the continuum emission of the shock heated, dense midplane.

@@

In this paper we have performed radiative transfer simulations of a hybrid model comprising a 0.4$\, M_\odot$ self gravitating disc with radius 64$\,AU$ showing spiral density waves, surrounded by an envelope simulated as a collapsing 10$\,M_\odot$ BE-sphere. CASA simulations show that at a distance of 100$\,$pc both spatial resolution of the spirals in such a disc and extraction of kinematic and temperature from molecular lines are possible in ALMA band 7. Our simulations show that many  molecular species are predominantly seen in absorption towards the centre of  self-gravitating protoplanetary discs. The quiescent nature of the envelope around such discs only obscures them within $\pm\,0.5 km\,s^{-1}$.\newline
One assumption made in this model is that the gas and dust are in thermal equilibrium.If they are not and the dust is significantly cooler than the gas then transitions may not show up in absorption.\newline
%The mass of the disc used (0.4$\,\rm{M}_\odot$) is on the high end for early discs (citations for estimate of early class 1 disc masses).\newline
In the model spirals shock heat the mid-plane layer and the colder more diffuse gas absorbs the continuum from it, this results in molecular lines being seen in absorbtion whenever lines of sight pass through the cold diffuse gas at larger heights on to the hot dense midplane of the disc. Rotation curves can be gathered from these observations even with spiral structure\newline
Outflows could contaminate measurements of rotation curves meaning that speices  which commonly trace outflows, such as CO and HCO$^+$ are not going to be good tracers of disc rotation. However speices such as OCS and C$^{18}$O which are not commonly seen in outflows (Yildiz et al 2012,)can be used to trace disc rotation.
